Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit

### 2016-07-05 paper/paper.tex

%\twocolumn[
%  \begin{@twocolumnfalse}
%\maketitle
%\hrule
\begin{abstract}
We consider two types of events, identified by decreases in $D_{st}$ below a threshold value and increases in the the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value on 1-hour and 1-day time scales using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events, we find that there is a weak and small amplitude ($< 10$\%) difference between $\rho_{eq}$ on the day of the event and the days before and after on both daily and hourly timescales.  Although very large geomagnetic storms have been observed to correspond to large increases in $\rho_{eq}$, on average, $\rho_{eq}$ following the onset of a geomagnetic storm does not depend on the storm intensity or north-south component of the interplanetary magnetic field, $B_z$. From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ six hours after the event onset corresponding to larger $\rho_{eq}$ events.  In contrast, the average of $B_z$ six hours prior to $\rho_{eq}$ events does not have a statistically significant impact on $\rho_{eq}$ after onset.
\end{abstract}
%\vspace{2em}
%\hrule
%\vspace{2em}
% \end{@twocolumnfalse}
%  ]

@@

The dependence of the plasmapause location on geomagnetic activity was observed as early as \cite{Carpenter1966} and there are well--validated models of the plasmapause boundary, as a function of $L$, $MLT$, and geomagnetic activity (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}).  Models also exist for the plasmasphere density (\cite{Gallagher1988EmpiricalModelPlasmasphere}; \cite{LemaireEarthsPlasmasphere}) with similar parameterization.

The dependence of the plasma density outside of the nominal plasmapause location on geomagnetic and solar wind conditions has not been extensively studied until the past decade (\cite{Takahashi2006}; \cite{Takahashi2010}; \cite{Denton2016}). Earlier works that considered profiles of hydrogen ion density found a significant dependence of the plasmasphere and plasmapause on geomagnetic activity (\cite{Chappell1970}; \cite{Maynard1977}; \cite{Carpenter1992}), but the relationship between geomagnetic activity and density outside of the plasmapause was not explicitly considered.

@@

\cite{Denton2006} found a weak relationship between $D_{st}$ and and $K_p$ and the field line distribution of mass density, $\rho$, using observations of the first three toroidal Alfv\'en harmonic frequencies for $L$ in the range of 6-8~$R_E$.  A more pronounced peak in the distribution along a filed line near the equator was found for lower $D_{st}$ and larger $K_p$.  The equatorial value of $\rho$ appeared identical for $D_{st}$ bins of [-142, -31] and [-31, 37] nT and $K_p$ bins of [1.5,3.4] and [3.4, 5.9].

@@

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincided with with significant changes in $\rho_{eq}$ for $L$ near 6.8~$R_E$. For five storms, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, the epoch average of $D_st$ exhibited for storms with a minimum $D_{st} < -50$~nT, an enhancement in $\rho_{eq}$ the same day as minimum $D_{st}$ epoch averages.

\cite{Yao2008} studied the relationship between $D_{st}$ and the number density of low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-shells (2-14) had a strong correlation (0.88) with the minimum $D_{st}$ of the storm.  Although the correlation at geosynchronous distances was not calculated, the data presented for four events show that the variations in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

The above results in the papers discussed indicate that (1) the mass density in the at geosynchronous distances is best correlated with $F10.7$ and (2) there exists a much weaker statistical relationship between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \cite{Denton2006} data set and attempt to identify and statistically characterize the relationship between geomagnetic activity with mass density at geosynchronous distances.  In addition, we attempt to identify solar wind processes that may drive changes in mass density.

@@

Figure~\ref{ccplot} shows the long-term trends of log$_{10}(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot} along with that for 1--day and 1--hour medians.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites is 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

\section{Epoch Average Results}

@@

Our first analysis uses data only from the time interval 1989-1991, which corresponds with the interval used in \cite{Takahashi2010}. In this interval, there were 329 $D_{st}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after each event.  To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across events.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for the 329 events from 1989-1991.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$ before and after the event, although the magnitude of increase observed here is $1.5$ amu/cm$^3$ (from 18 to 19.5~amu/cm$^3$) instead of the increase of $\sim 10$ amu/cm$^3$ found in Figure 11 of \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and red error lines.  Error lines for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values. For the non-$\rho_{eq}$\ medians, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of $\rho_{eq}$\ values as all available measurements were used instead of restricting to only values where a $\rho_{eq}$ value also existed.

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6 - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.

@@

These tests indicate that there is not a statistically significant change in medians of $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.  In addition, the observed differences are on the order of only 10\% of the $\rho_{eq}$, which is small compared to the variance in the observations.

\begin{table}
\small
%\input{../tables/DeltaBootstraps-case10.txt}
\caption{Results of test on of means of $\rho_{eq}$ shown in the top panel of Figure~\ref{DailyAveragedDstEvents} between days of threshold crossing  near (day = 1 or -1) or on the day of a $D_{st}$ event (day = 0).}
\label{BootstrapDifferenceTable}
\end{table}

@@

The bottom panel of Figure~\ref{HourlyAveragedDstEvents} contains the events shown in the top panel under the constraint that $D_{st}$ remained below -50~nT for at least twelve hours after the event onset time.  In the time period of elevated mass density within 12 hours of the start of the event, there are only $\sim$5 samples per hourly interval.  This result is consistent with previous observations of large mass density spikes in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

@@

\subsection{Conclusions}

@@

\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedPdyn-case24-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedPdyn-case24-t025-tf30-GOES6.eps}
  \caption{Binning $\rho_{eq}$ events by $B_z$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:PdynBinned}
\end{figure}

\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedV_SW-case24-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedV_SW-case24-t025-tf30-GOES6.eps}
  \caption{Binning $\rho_{eq}$ events by $B_z$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:V_SWBinned}
\end{figure}

\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedkp-case24-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinnedkp-case24-t025-tf30-GOES6.eps}
  \caption{Binning $\rho_{eq}$ events by $B_z$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:KpBinned}
\end{figure}

\begin{figure}[tp!]
  \centering
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinneddst-case24-t020-tf25-GOES6.eps}
  \includegraphics[scale=0.40]{figures/RhoBinned/RhoBinneddst-case24-t025-tf30-GOES6.eps}
  \caption{Binning $\rho_{eq}$ events by $B_z$ value (a) at onset and four hours before, and (b) at onset and four hours after.}
  \label{fig:DstBinned}
\end{figure}

>>>>

%\twocolumn[
%  \begin{@twocolumnfalse}
%\maketitle
%\hrule
\begin{abstract}
We consider two types of events, identified by decreases in $D_{st}$ below a threshold value and increases in the the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value on 1-hour and 1-day time scales using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events, we find that there is a weak and small amplitude ($< 10$\%) difference between $\rho_{eq}$ on the day of the event and the days before and after on both daily and hourly timescales.  Although very large geomagnetic storms have been observed to correspond to large increases in $\rho_{eq}$, on average, $\rho_{eq}$ following the onset of a geomagnetic storm does not depend on the storm intensity or the north-south component of the interplanetary magnetic field, $B_z$. From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ six hours after the event onset corresponding to larger $\rho_{eq}$ events.  In contrast, the average of $B_z$ six hours prior to $\rho_{eq}$ events does not have a statistically significant impact on $\rho_{eq}$ after onset.
\end{abstract}
%\vspace{2em}
%\hrule
%\vspace{2em}
% \end{@twocolumnfalse}
%  ]

@@

The dependence of the plasmapause location on geomagnetic activity was observed as early as \cite{Carpenter1966} and there are well--validated models of the plasmapause boundary as a function of $L$, $MLT$, and geomagnetic activity (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}).  Models also exist for the plasmasphere density (\cite{Gallagher1988EmpiricalModelPlasmasphere}; \cite{LemaireEarthsPlasmasphere}) with similar parameterization.

The dependence of the plasma density outside of the nominal plasmapause location on geomagnetic and solar wind conditions has not been extensively studied until the past decade (\cite{Takahashi2006}; \cite{Takahashi2010}; \cite{Denton2016}). Earlier works that considered profiles of hydrogen ion density found a significant dependence plasmapause location on geomagnetic activity (\cite{Chappell1970}; \cite{Maynard1977}; \cite{Carpenter1992}), but the relationship between geomagnetic activity and density outside of the plasmapause was not explicitly considered.

@@

\cite{Denton2006} found a weak relationship between $D_{st}$ and and $K_p$ and the field line distribution of mass density, $\rho$, using observations of the first three toroidal Alfv\'en harmonic frequencies for $L$ in the range of 6-8~$R_E$.  A more pronounced peak in the distribution along a field line near the equator was found for lower $D_{st}$ and larger $K_p$.  The equatorial value of $\rho$ were found to be similar for $D_{st}$ bins of [-142, -31] and [-31, 37] nT and $K_p$ bins of [1.5,3.4] and [3.4, 5.9].

@@

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincided with with significant changes in $\rho_{eq}$ for $L$ near 6.8~$R_E$. For five storms, two had $\rho_{eq}$ spikes after $D_{st}$ minimum, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, the epoch average of $D_{st}$ exhibited for storms with a minimum $D_{st} < -50$~nT showed an enhancement in $\rho_{eq}$ the same day as minimum $D_{st}$ in the epoch averages.

\cite{Yao2008} studied the relationship between $D_{st}$ and the number density of low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-shells (2-14) had a strong correlation (0.88) with the minimum $D_{st}$ of the storm.  Although the correlation at geosynchronous distances was not calculated, the data presented for four events show that the enhancements in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

The above results indicate that (1) the mass density in the at geosynchronous distances is best correlated with $F10.7$ and (2) there exists a much weaker statistical relationship between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \cite{Takahashi2010} data set and attempt to identify and statistically characterize the relationship between geomagnetic activity with mass density at geosynchronous distances.  In addition, we attempt to identify solar wind processes that may drive changes in mass density.

@@

Figure~\ref{ccplot} shows the long-term trends of log$_{10}(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot} along with that for 1--day and 1--hour medians.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used combined measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites is 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29, respectively.

\section{Results}

@@

Our first analysis uses GOES-6 data only from the time interval 1989-1991, which corresponds with the interval used in \cite{Takahashi2010}. In this interval, there were 329 $D_{st}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after event onset.  To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across events in each epoch day.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for the 329 events from 1989-1991.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$ before and after the event, although the magnitude of increase observed here is $1.5$ amu/cm$^3$ (from 18 to 19.5~amu/cm$^3$) instead of the increase of $\sim 10$ amu/cm$^3$ found in Figure 11 of \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and red error lines.  Error lines for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values. For the non-$\rho_{eq}$\ parameters, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of $\rho_{eq}$\ values as all available measurements were used instead of restricting to only values where a $\rho_{eq}$ value also existed.

In the lower panel of Figure \ref{DailyAveragedDstEvents}, it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6 - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.

@@

These tests indicate that there is not a statistically significant difference in medians of $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.  In addition, the observed differences are on the order of only 10\% of the local average of $\rho_{eq}$, which is small compared to the variance in the observations.

\begin{table}
\small
\centering
\input{../tables/DeltaBootstraps-case10.txt}
\caption{Results of test on of means of $\rho_{eq}$ shown in the top panel of Figure~\ref{DailyAveragedDstEvents} between days of threshold crossing  near (day = 1 or -1) or on the day of a $D_{st}$ event (day = 0).}
\label{BootstrapDifferenceTable}
\end{table}

@@

The bottom panel of Figure~\ref{HourlyAveragedDstEvents} contains the events shown in the top panel under the constraint that $D_{st}$ remained below -50~nT for at least twelve hours after the event onset time.  In the time period of elevated mass density within 12 hours of the start of the event, there are only $\sim$4 samples per hourly interval.  This result is consistent with previous observations of large mass density spikes in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

@@

\subsection{Summary and Conclusions}

In recent years, various works have considered the influence of solar wind control and the geomagnetic activity relationship to the equatorial plasma mass density at geosynchronous orbit, generally using event studies and linear correlations (\cite{Takahashi2006}; \cite{Denton2006}; \cite{Yao2008}; \cite{Takahashi2010}; \cite{Denton2016}).  The event studies show a possible relationship between $\rho_{eq}$ and solar wind and geomagnetic activity indicators such as $K_p$ and $D_{st}$ while correlation studies have generally found a very weak relationship.

This work presents an analysis of two types of events intended to provide insight into the relationship between $\rho_{eq}$ and solar wind and geomagnetic parameters by considering both large geomagnetic events and large $\rho_{eq}$ events.  The analysis attempts to address a major difficulty with the identification of the cause in $\rho_{eq}$ enhancements, which can occur for two reasons: (1) an extended period of positive $B_z$, which leads to low geomagnetic activity, plasmasphere refilling, and a possible expansion of the plasmasphere to geosynchronous altitudes and (2) strong negative $B_z$, which (a) leads to the inward motion of the higher density plasmasphere and leaves geosynchronous altitudes in the lower density plasma trough just outside of the plasmapause and (b) causes possible magnetospheric processes to enhance the mass density at geosynchronous altitudes.

On average, around the time of large geomagnetic events, there is a very weak enhancement of $\rho_{eq}$ and this enhancement does not depend on either the interplanetary magnetic field $B_z$ or the intensity of the geomagnetic event.  This indicates that there is an approximate balance between magnetospheric enhancement in $\rho_{eq}$ and any reduction due to the inward motion of the plasmasphere.

$\rho_{eq}$ events are larger when $B_{z}$ is more positive after the onset of the event, which indicates that a substantial portion of the events may due to the apparent mass refilling mechanism described by \cite{Denton2016}.  Without additional information about the location of the plasmapause, it may not be possible to separate out the conditions required for enhancement due to enhanced magnetospheric activity from this mechanism.

@@
